Bayesian assessment of heterogeneity in psychological phenomena

### 2023-09-07 ms.Rmd
Incoming revisions

---
title: 'Bayesian assessment of heterogeneity in psychological phenomena'
shorttitle: Heterogeneity in psychology
leftheader: Heterogeneity in psychology
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Tilburg School of Social and Behavioral Sciences, Tilburg University
    email: m.j.vuorre@tilburguniversity.edu
  - name: Matthew Kay
    affiliation: 2
  - name: Niall Bolger
    affiliation: 3
affiliation:
  - id: 1
    institution: Department of Social Psychology, Tilburg University
  - id: 2
    institution: Computer Science and Communication Studies, Northwestern University
  - id: 3
    institution: Department of Psychology, Columbia University
abstract: |
  Advances in statistics and data collection methods have brought variability in psychological phenomena to the forefront of theoretical development. Yet, established methods and practices for calculating and describing heterogeneity are lacking in how they treat uncertainty regarding heterogeneity in the phenomenon under study. In the current work, we aim to deliver methods and guidelines for more complete treatments of heterogeneity in psychological phenomena.
  We provide a walkthrough of probabilistic methods for i. estimating, ii. summarising, and iii. describing heterogeneity. The methods we propose---bayesian hierarchical models---are especially suitable for investigating heterogeneity because they naturally describe uncertainty in all model parameters.
  We hope that the proposed methods help researchers to move from reporting point estimates to drawing inferences about magnitudes and differences in heterogeneity in psychological phenomena across different populations.
keywords: variation, heterogeneity, hierarchical models, mixed models, methodology
authornote: |
  \noindent \textbf{Working paper. Proceed with caution.}
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: references.bib
floatsintext: yes
numbersections: false
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
header-includes:
  - |
    \usepackage{float}
    \setlength{\parskip}{4pt}
output: 
  papaja::apa6_pdf:
    number_sections: true
    keep_tex: false
    highlight: kate
  papaja::apa6_docx: 
    number_sections: true
    keep_tex: false
---

@@

```

```{r data, include = FALSE}
# Load, save, & clean example data from Bolger et al. (2019)
dir.create("data", FALSE)
path <- "data/bolger-etal.zip"
if (!file.exists(path)) {
  download.file(
    "https://github.com/kzee/heterogeneityproject/archive/refs/heads/master.zip",
    destfile = path
  )
  unzip(path, exdir = dirname(path), junkpaths = TRUE)
}

# Save variables as numerical & factor (might need for more complex contrast/eff/etc codes)
dat <- read_csv("data/heterogeneity_dataset1_traitvalence.csv") |> 
  # Cleaning as in Bolger et al 2019
  filter(
    response.keys == "up",
    scale(rt) < 3,
    !(id %in% c(250, 257, 272))
  ) |> 
  # Some more cleaning
  mutate(
    person = fct_anon(factor(id)),
    trial,
    valence = factor(valenceE, levels = c(-0.5, 0.5), labels = c("Negative", "Positive")),
    rt = logrt
  ) |> 
  select(person, trial, valence, rt) |> 
  arrange(person, trial)
```

@@

One reason for the scarcity of reporting and interpreting heterogeneity is that psychologists still tend to analyze data with models that cannot address between-person variability, such as ANOVA [@bolgerCausalProcessesPsychology2019]. Although more appropriate mixed-effects (or multilevel, hierarchical) models are rapidly becoming more widespread, we believe a second important reason is that researchers do not yet have the conceptual and practical tools required to estimate and describe heterogeneity to a sufficiently informative degree. In the current work, we propose several practical methods, descriptions, and visualizations that allow researchers to better estimate, interpret, and report heterogeneity in a manner that maximizes empirical results' impact on theory building.

@@

@@

```{r}
#| label: tab-dat1
dat |> 
  head() |> 
  apa_table(
    span_text_columns = FALSE, 
    digits = c(0,0,0,2),
    caption = "First six rows of example dataset 1.", 
    note = "rt is log-transformed."
  )
```

@@

This equation includes two kinds of parameters: $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V), do not have subscripts. That means that they are not further modelled, and therefore sometimes called "fixed". We contrast coded valence such that negative words are coded as -0.5, and positive words as 0.5. This coding results in an intercept that corresponds to the average reaction time across negative and positive words, and a slope coefficient that reflects the difference in log(rt) between negative and positive words. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$ are subject-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the average reaction time for subject $j=1$. The $\gamma$ coefficients have subscripts, which in this context indicates that they are further modelled, and therefore sometimes called "random" [@gelmanDataAnalysisUsing2007]. As is standard in multilevel modelling, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

@@

tab1_lmer |> 
  mutate(
    across(where(is.numeric), ~number(., .01)),
    MLE = str_glue("{Coefficient} [{CI_low}, {CI_high}]"),
  ) |> 
  select(Parameter, `Estimate [95\\%CI]` = MLE) |> 
  mutate(
    Parameter = m1_parnames
  ) |> 
  apa_table(
    span_text_columns = FALSE,
    caption = "Parameter estimates from Model 1.",
    escape = FALSE
  )
mu1n <- tab1_lmer[2,2,T]
mu1 <- number(mu1n, .01)
sd1n <- tab1_lmer[4,2,T]
sd1 <- number(sd1n, .01)
```

We show a traditional summary of the estimated parameters from this model in Table \@ref(tab:tbl-fit-1). The expected effect of positive valence on RT for the average person is approximately `r mu1` log seconds, and the expected standard deviation describing this effect's degree of variability in the population is approximately `r sd1`.

@@

```{r}
#| label: hdist-1
#| fig.height: 2.2
#| fig.width: 8
#| fig.env: "figure*"
#| fig.cap: The heterogeneity distribution and various descriptions of the expected heterogeneity in valence effects from Model 1. \textbf{A}. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. \textbf{B}. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. \textbf{C}. Proportion of negative valence effects in the population (dark). \textbf{D}. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

@@

While the interval and ratio descriptors allow determining where the population's slope values are likely to fall, they do so in absolute terms, such as log(rt) in the running example. One way to contextualize the magnitude of heterogeneity is to put it in relative terms. To do so, we can compare the standard deviation of the valence effect distribution to its mean. [@bolgerCausalProcessesPsychology2019, p.609] suggest as a rule of thumb that if the ratio of the standard deviation to the average effect is 0.25 or greater, then heterogeneity can be deemed noteworthy. With these data and model, the ratio $\frac{\tau_1}{\beta_1}$ is `r ratio`, suggesting that the degree of heterogeneity in valence effects is noteworthy. (We reiterate the point made by @bolgerCausalProcessesPsychology2019 that the cutoff value of 0.25 is arbitrary and researchers should choose the cutoff based on their substantive goals.)

@@

Armed with the bayesian estimates (Table \@ref(tab:tbl-samples-1)), we now return to the distribution of valence effects in the population. We first redraw the expected heterogeneity distribution from Figure \@ref(fig:hdist-1) using the posterior mean values of $\beta_1$ and $\tau_1$ in Figure \@ref(fig:hdist-2)A. Superimposed to that normal density curve are heterogeneity distributions calculated not from the posterior means, but from 100 random samples from the posterior distribution. From these curves we glean that the distribution of valence effects might well be narrower (i.e. less heterogeneity) or wider (more heterogeneity) than is suggested by the point estimates.

Moreover, the location of these curves differ: Some curves are further to the left (average effect is more negative), and some further to the right (average effect is more positive). The distribution of these curves represents our uncertainty over the heterogeneity distribution of valence effects in the population. Thus, a good summary of heterogeneity that incorporates uncertainty must include information about uncertainty in both the scale (standard deviation) and location (mean) of the heterogeneity distribution.

```{r}
#| label: hdist-2
#| fig.height: 3.4
#| fig.width: 4
#| fig-cap: Posterior distribution of the heterogeneity distribution of valence effects (\textbf{A}) and its 90\% heterogeneity interval (\textbf{B}). \textbf{A}. Posterior mean heterogeneity distribution in dark blue, and 100 random draws from the distribution's posterior distribution in light blue. \textbf{B}. 100 random draws from the 90\% heterogeneity interval's posterior distribution.

@@

To carry uncertainty forward from model parameters to their transformations, such as $HI_{90}$, we reproduce the calculations described above, but instead of using the point estimates we repeat the calculation over each posterior sample. In this manner, we obtain (samples from) posterior distributions of the parameter transformations, such as $HI_{90}$ (Figure \@ref(fig:hdist-2)B).

@@

```{r}
#| label: hdist-3
#| fig.height: 2.2
#| fig.width: 4
#| fig-cap: \textbf{A}. 100 random draws from the valence effect distribution's posterior distribution. Darker area under the curve represents the proportion of negative effects in the population for that posterior draw ($p^d$). \textbf{B}. Histogram of all 4,000 draws from the posterior distribution of $p^d$, and its posterior mean and 95%CI.

@@

However, a sign test against zero is not always the most informative quantity because it discards all information about the magnitude of effects. A more theoretically motivated approach is to argue for and define a smallest effect size of interest (SESOI) which then defines a region of practical equivalence (ROPE) to zero, and then quantify uncertainty for the proposition that the effect is "practically equivalent to zero". Figure \@ref(fig:hdist-4)A shows 100 random draws of the heterogeneity distribution of valence effects, such that the area under the curve within a ROPE of [-0.1, 0.1] ($p^{ROPE}$) is filled with a darker color. That shaded area represents the posterior probability, for each draw, that the effect is practically equivalent to zero. Figure \@ref(fig:hdist-4)B then shows a histogram of all 4,000 posterior draws of $p^{ROPE}$ with their mean and 95%CI. We learn that `r tmp$rope_r` of individuals in the population have a valence effect that is practically equivalent to zero.

```{r}
#| label: hdist-4
#| fig.height: 2.2
#| fig.width: 4
#| fig-cap: \textbf{A}. 100 random draws from the valence effect distribution's posterior distribution. Dark area under the curve represents the proportion of effects within a region of practical equivalence to zero (ROPE) in the population for that posterior draw ($p^{ROPE}$). \textbf{B}. Histogram of all 4,000 draws from the posterior distribution of $p^{ROPE}$, and its posterior mean and 95%CI.
p1 <- distinct(dist_samples, Sample, b_valence1, sd_person__valence1) |> 
  arrange(b_valence1) |> 
  # slice(seq(1, nrow(.), by = 2)) |>
  mutate(i = 1:n()) |> 
  mutate(
    dist = dist_normal(b_valence1, sd_person__valence1)
  ) |> 
  ggplot() +
  scale_fill_manual(
    values = c("dodgerblue1", "dodgerblue3"), 
    na.value = "white"
    ) +
  scale_x_continuous(
    "Valence effect",
    expand = expansion(0),
    breaks = extended_breaks()
  ) +
  scale_y_continuous(
    "Sample",
    expand = expansion(0)
  ) +
  aes(xdist = dist, y = i) +
  geom_vline(xintercept = 0, linewidth = .2, linetype = "dashed") +
  stat_slab(
    aes(fill = after_stat(near(x, 0, .1))), 
    height = 15,
    slab_color = "grey10",
    slab_linewidth = .1,
    alpha = .5,
    p_limits = c(.01, .99),
    show.legend = FALSE
  )
p2 <- post1 |> 
  mutate(
    dist = dist_normal(b_valence1, sd_person__valence1),
    pd = cdf(dist, 0),
    prope = cdf(dist, 0.1) - cdf(dist, -0.1)
  ) |> 
  ggplot(aes(prope)) +
  scale_y_continuous(
    "Count",
    expand = expansion(0)
  ) +
  scale_x_continuous(
    TeX('$p^{ROPE}$'),
    breaks = extended_breaks(),
    labels = ~percent(., 1)
  ) +
  stat_histinterval(
    breaks = 50,
    outline_bars = TRUE,
    slab_color = "gray10",
    slab_fill = alpha("dodgerblue1", .33),
    slab_linewidth = .2,
    justification = .01,
    point_color = "dodgerblue4", 
    interval_color = "dodgerblue4",
    .width = .90
  )
(p1 | p2) &
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  ) &
  plot_annotation(tag_levels = "A")
```

So far, these examples have highlighted the importance of quantifying uncertainty in measures of heterogeneity. Had we only focused on the point estimates (posterior means), we might have misleadingly concluded that $p^d$=`r percent(tmp$pd, .1)` and $p^{ROPE}$=`r percent(tmp$rope, .1)`. However, with 95% confidence, these values might be as great as `r percent(tmp$pd.upper, .1)` and `r percent(tmp$rope.upper, .1)`, respectively.

@@

```{r}
#| label: hdist-5
#| fig.height: 2.2
#| fig.width: 4
#| fig-cap: \textbf{A}. 4,000 random draws from the posterior distribution of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). \textbf{B}. Histogram of 4,000 draws from the posterior distribution of the ratio of the valence distribution's scale over its  location $\frac{\tau_1}{\beta_1}$, and its posterior mean and 95%CI.
p1 <- post1 |> 
  ggplot(aes(b_valence1, sd_person__valence1)) +
  scale_y_continuous(
    TeX("$\\tau_1$")
  ) +
  scale_x_continuous(
    TeX("$\\beta_1$")
  ) +
  geom_point(
    shape = 1,
    col = "dodgerblue3",
    alpha = .25,
    size = .5
  )
p2 <- post1 |> 
  mutate(ratio = abs(sd_person__valence1 / b_valence1)) |> 
  ggplot(aes(ratio)) +
    scale_y_continuous(
    "Count",
    expand = expansion(0)
  ) +
  scale_x_continuous(
    TeX('$\\frac{\\tau_1}{\\beta_1}$'),
    breaks = extended_breaks()
  ) +
  stat_histinterval(
    breaks = 50,
    outline_bars = TRUE,
    slab_color = "gray10",
    slab_fill = alpha("dodgerblue1", .33),
    slab_linewidth = .2,
    justification = .01,
    point_color = "dodgerblue4", 
    interval_color = "dodgerblue4",
    .width = .90
  ) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
  )
(p1 | p2) &
  theme(
    aspect.ratio = 1
  ) &
  plot_annotation(tag_levels = "A")
```

>>>>

---
title: 'Bayesian assessment of heterogeneity in psychological phenomena'
shorttitle: Heterogeneity in psychology
leftheader: Heterogeneity in psychology
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Tilburg School of Social and Behavioral Sciences, Tilburg University
    email: m.j.vuorre@tilburguniversity.edu
  - name: Matthew Kay
    affiliation: 2
  - name: Niall Bolger
    affiliation: 3
affiliation:
  - id: 1
    institution: Department of Social Psychology, Tilburg University
  - id: 2
    institution: Computer Science and Communication Studies, Northwestern University
  - id: 3
    institution: Department of Psychology, Columbia University
abstract: |
  Advances in statistics and data collection methods have brought variability in psychological phenomena to the forefront of theoretical development. Yet, established methods and practices for describing heterogeneity underestimate the true uncertainty in those estimates. In the current work, we aim to deliver methods and guidelines for more rigorous treatments of heterogeneity in psychological phenomena.
  We provide a walkthrough of methods for i. estimating, ii. summarising, and iii. describing heterogeneity. Although other methods for incorporating estimation uncertainty are available, the methods we propose---involving bayesian hierarchical models---are especially suitable for investigating heterogeneity because they naturally describe uncertainty in all model parameters.
  We hope that the proposed methods help researchers to widen their inferential focus from statistical averages to include statistical heterogeneity in psychological phenomena.
keywords: variation, heterogeneity, hierarchical models, mixed models, methodology
authornote: |
  \noindent \textbf{This working paper is not yet peer reviewed.}
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: references.bib
floatsintext: yes
numbersections: false
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
header-includes:
  - |
    \usepackage{float}
    \setlength{\parskip}{4pt}
output: 
  papaja::apa6_pdf:
    number_sections: true
    keep_tex: false
    highlight: kate
  papaja::apa6_docx: 
    number_sections: true
    keep_tex: false
---

@@

dir.create("data", FALSE)
```

@@

One reason for the scarcity of reporting and interpreting heterogeneity is that psychologists still tend to analyze data with models that obscure the assessment of heterogeneity, such as ANOVA [@bolgerCausalProcessesPsychology2019]. Although more appropriate hierarchical (or multilevel, mixed-effects) models are rapidly becoming more widespread, we believe a second important reason is that researchers do not yet have the conceptual and practical tools required to estimate and describe heterogeneity. In the current work, we propose several practical methods, descriptions, and visualizations that allow researchers to better estimate, interpret, and report heterogeneity in a manner that maximizes empirical results' impact on theory building.

@@

```{r data-bolgeretal, include = FALSE}
# Load, save, & clean example data from Bolger et al. (2019)
path <- "data/bolger-etal.zip"
if (!file.exists(path)) {
  download.file(
    "https://github.com/kzee/heterogeneityproject/archive/refs/heads/master.zip",
    destfile = path
  )
  unzip(path, exdir = str_remove(path, ".zip"), junkpaths = TRUE)
}

# Save variables as numerical & factor (might need for more complex contrast/eff/etc codes)
dat <- read_csv("data/bolger-etal/heterogeneity_dataset1_traitvalence.csv") |> 
  # Cleaning as in Bolger et al 2019
  filter(
    response.keys == "up",
    scale(rt) < 3,
    !(id %in% c(250, 257, 272))
  ) |> 
  # Some more cleaning
  mutate(
    person = fct_anon(factor(id)),
    trial,
    valence = factor(valenceE, levels = c(-0.5, 0.5), labels = c("Negative", "Positive")),
    rt = logrt
  ) |> 
  select(person, trial, valence, rt) |> 
  arrange(person, trial)
```

@@

```{r}
#| label: tab-dat1
dat |> 
  head() |> 
  apa_table(
    span_text_columns = FALSE, 
    digits = c(0,0,0,2),
    caption = "First six rows of example dataset 1 (Bolger et al., 2019; Exp 1).", 
    note = "rt is log-transformed."
  )
```

@@

This equation includes two kinds of parameters: $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V), do not have subscripts. That means that they are not further modelled except by way of (optional) prior distributions with fixed hyperparameters. In line with existing conventions in bayesian inference and software, we refer to these as *population level* parameters. From the frequentist perspective, these terms are constants (i.e. do not potentially vary as functions of covariates) and are therefore sometimes "fixed". For all practical purposes, because we do not discuss prior distributions with constant hyperparameters, readers can equate the terms "fixed" and "population level" parameters in the current discussion.

We contrast coded valence such that negative words are coded as -0.5, and positive words as 0.5. This coding results in an intercept that corresponds to the average reaction time across negative and positive words, and a slope coefficient that reflects the difference in log(rt) between negative and positive words. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$ are subject-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the average reaction time for subject $j=1$. The $\gamma$ coefficients have subscripts, which in this context indicates that they are further modelled on covariates---here, index variables for "person" in the data. We refer to these as "group level" parameters, in line with conventions. In the frequentist discourse, these parameters are called "random" parameters [@gelmanDataAnalysisUsing2007]. Again, readers can equate the term "group level" parameters with "random" parameters in the ongoing discussion.

As is standard in multilevel modelling, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

@@

tab1_lmer |> 
  mutate(
    across(where(is.numeric), ~number(., .01)),
    MLE = str_glue("{Coefficient} [{CI_low}, {CI_high}]"),
  ) |> 
  select(Parameter, `Estimate [95\\%CI]` = MLE) |> 
  mutate(
    Parameter = m1_parnames
  ) |> 
  apa_table(
    span_text_columns = FALSE,
    caption = "Parameter estimates from Model 1.",
    escape = FALSE
  )
mu1 <- number(as.numeric(tab1_lmer[2,2:4,T]))
sd1 <- number(as.numeric(tab1_lmer[4,2:4,T]))
mu1n <- tab1_lmer[2,2,TRUE]
sd1n <- tab1_lmer[4,2,TRUE]
```

We show a traditional summary of the estimated parameters from this model in Table \@ref(tab:tbl-fit-1). For the average person, the estimated effect of valence on RT is `r mu1[1]` log seconds, with a 95% bootstrap confidence interval (CI) extending from `r mu1[2]` to `r mu1[3]`. The estimated standard deviation of between-person differences in the valence effects is `r sd1[1]` log seconds, with a 95% CI of [`r sd1[2]`, `r sd1[3]`].

@@

```{r}
#| label: hdist-1
#| fig.height: 2.2
#| fig.width: 8
#| fig.env: "figure*"
#| fig.cap: The heterogeneity distribution and various descriptions of the expected heterogeneity in valence effects from Model 1. **A**. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. **B**. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. **C**. Proportion of negative valence effects in the population (dark). **D**. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

@@

Although the interval and proportion descriptors show where the population's slope values are likely to fall, they do so in absolute terms, such as log(rt) in the running example. A contrasting, relative way to show heterogeneity is to express it as a ratio of the standard deviation to the mean. In our example this is the ratio of the standard deviation to the mean of the valence effect.

[@bolgerCausalProcessesPsychology2019, p.609] suggest as a rule of thumb that if the ratio of the standard deviation to the average effect is 0.25 or greater, then heterogeneity can be deemed noteworthy. With these data and model, the ratio $\frac{\tau_1}{\beta_1}$ is `r ratio`, suggesting that the degree of heterogeneity in valence effects is noteworthy. (We reiterate the point made by @bolgerCausalProcessesPsychology2019 that the cutoff value of 0.25 is arbitrary and researchers should choose the cutoff based on their substantive goals.)

@@

Armed with the bayesian estimates (Table \@ref(tab:tbl-samples-1)), we now return to the distribution of valence effects in the population. We first redraw the expected heterogeneity distribution from Figure \@ref(fig:hdist-1) using the posterior mean values of $\beta_1$ and $\tau_1$ in Figure \@ref(fig:hdist-2)A. Superimposed on that normal density curve are heterogeneity distributions calculated not from the posterior means, but from 100 randomly sampled values from the joint posterior distribution of $\beta_1$ and $\tau_1$. From these curves we glean that the distribution of valence effects might well be narrower (i.e. less heterogeneity) or wider (more heterogeneity) than is suggested by the point estimates.

Moreover, the location of these curves differ: Some curves are further to the left (average effect is more negative), and some further to the right (average effect is more positive). The distribution of these curves represents our uncertainty over the heterogeneity distribution of valence effects in the population. Thus, a good summary of heterogeneity that incorporates uncertainty must include information about uncertainty in both the location (mean) and scale (standard deviation) parameters of the heterogeneity distribution.

```{r}
set.seed(99)
tmp <- post1 |> 
  slice_sample(n = 4) |> 
  mutate(dist = dist_normal(b_valence1, sd_person__valence1)) |> 
  mutate(i = 1:n())

tmp <- tmp |> 
  mutate(
    plot = map2(
      b_valence1, sd_person__valence1, 
      ~tibble(dist = dist_normal(.x, .y)) |> 
        ggplot(aes(xdist = dist)) +
        scale_x_continuous(
          breaks = extended_breaks(3)
        ) +
        coord_cartesian(xlim = c(-0.4, 0.1), clip = "off") +
        scale_y_continuous(
          expand = expansion(c(0.0, 0.1))
        ) +
        stat_slab(slab_fill = "grey30") +
        theme(
          axis.title = element_blank(),
          axis.ticks.x = element_line(size = rel(0.33)),
          axis.ticks.y = element_blank(),
          axis.text.y = element_blank(),
          axis.text.x = element_text(size = rel(0.6)),
          axis.line.y = element_blank()
          )
      )
  )

p1 <- post1 |> 
  ggplot(aes(b_valence1, sd_person__valence1)) +
  scale_x_continuous(
    expand = expansion(.33)
  ) +
  scale_y_continuous(
    expand = expansion(.33)
  ) +
  geom_point(shape = 1, col = "dodgerblue") +
  geom_curve(
    data = tmp |> 
      mutate(y = sd_person__valence1),
    aes(x = b_valence1, xend = 0,
        y = y, yend = y),
    alpha = .33, linewidth = .75, curvature = .2
  ) +
  theme(aspect.ratio = 1)

p1 + 
  inset_element(tmp$plot[[1]], 0.7, 0.1, 0.3, 0.28)
```

```{r}
#| label: hdist-2
#| fig.height: 3.4
#| fig.width: 4
#| fig-cap: Posterior distribution of the heterogeneity distribution of valence effects (**A**) and its 90\% heterogeneity interval (**B**). **A**. Posterior mean heterogeneity distribution in dark blue, and 100 random draws from the distribution's posterior distribution in light blue. **B**. 100 random draws from the 90\% heterogeneity interval's posterior distribution.

@@

To carry uncertainty forward from model parameters to their transformations, such as $HI_{90}$, we reproduce the calculations described above, but instead of using the typical (expected) values of the mean and standard deviation, we repeat the calculation for each of the 4,000 randomly sampled pairs of $\beta_1$ and $\tau_1$. In this manner, we get (samples from) posterior distributions of the parameter transformations, such as $HI_{90}$ (Figure \@ref(fig:hdist-2)B).

@@

```{r}
#| label: hdist-3
#| fig.height: 2.2
#| fig.width: 4
#| fig-cap: **A**. 100 random draws from the valence effect distribution's posterior distribution. Darker area under the curve represents the proportion of negative effects in the population for that posterior draw ($p^d$). **B**. Histogram of all 4,000 draws from the posterior distribution of $p^d$, and its posterior mean and 95%CI.

@@

However, a sign test against zero is not always the most informative quantity because it discards all information about the magnitude of effects. A more theoretically motivated approach is to define a smallest effect size of interest (SESOI) which then defines a region of practical equivalence (ROPE) where the effect is deemed "practically equivalent to zero". Figure \@ref(fig:hdist-4)A shows 100 random draws of the heterogeneity distribution of valence effects, such that the area under the curve within a ROPE of [-0.1, 0.1] ($p^{ROPE}$) is filled with a darker color. That shaded area under the curve represents the posterior probability, for each random sample from the join posterior distribution of $\beta_1$ and $\tau_1$, that the effect is practically equivalent to zero. Figure \@ref(fig:hdist-4)B then shows a histogram of all 4,000 posterior draws of $p^{ROPE}$ with their mean and 95%CI. We learn that `r tmp$rope_r` of individuals in the population have a valence effect that is practically equivalent to zero.

```{r}
#| label: hdist-4
#| fig.height: 2.2
#| fig.width: 4
#| fig-cap: **A**. 100 random draws from the valence effect distribution's posterior distribution. Dark area under the curve represents the proportion of effects within a region of practical equivalence to zero (ROPE) in the population for that posterior draw ($p^{ROPE}$). **B**. Histogram of all 4,000 draws from the posterior distribution of $p^{ROPE}$, and its posterior mean and 95%CI.
p1 <- distinct(dist_samples, Sample, b_valence1, sd_person__valence1) |> 
  arrange(b_valence1) |> 
  # slice(seq(1, nrow(.), by = 2)) |>
  mutate(i = 1:n()) |> 
  mutate(
    dist = dist_normal(b_valence1, sd_person__valence1)
  ) |> 
  ggplot() +
  scale_fill_manual(
    values = c("dodgerblue1", "dodgerblue4"), 
    na.value = "white"
    ) +
  scale_x_continuous(
    "Valence effect",
    expand = expansion(0),
    breaks = extended_breaks()
  ) +
  scale_y_continuous(
    "Sample",
    expand = expansion(0)
  ) +
  aes(xdist = dist, y = i) +
  geom_vline(xintercept = 0, linewidth = .2, linetype = "dashed") +
  stat_slab(
    aes(fill = after_stat(near(x, 0, .1))), 
    height = 15,
    slab_color = "grey10",
    slab_linewidth = .1,
    alpha = .5,
    p_limits = c(.01, .99),
    show.legend = FALSE
  )
p2 <- post1 |> 
  mutate(
    dist = dist_normal(b_valence1, sd_person__valence1),
    pd = cdf(dist, 0),
    prope = cdf(dist, 0.1) - cdf(dist, -0.1)
  ) |> 
  ggplot(aes(prope)) +
  scale_y_continuous(
    "Count",
    expand = expansion(0)
  ) +
  scale_x_continuous(
    TeX('$p^{ROPE}$'),
    breaks = extended_breaks(),
    labels = ~percent(., 1)
  ) +
  stat_histinterval(
    breaks = 50,
    outline_bars = TRUE,
    slab_color = "gray10",
    slab_fill = alpha("dodgerblue1", .33),
    slab_linewidth = .2,
    justification = .01,
    point_color = "dodgerblue4", 
    interval_color = "dodgerblue4",
    .width = .90
  )
(p1 | p2) &
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  ) &
  plot_annotation(tag_levels = "A")
```

So far, these examples have highlighted the importance of quantifying uncertainty in measures of heterogeneity. Had we only focused on the point estimates (posterior means), we might have misleadingly concluded that $p^d$ = `r percent(tmp$pd, .1)` and $p^{ROPE}$ = `r percent(tmp$rope, .1)`. However, with 95% confidence, these values might be as small as `r percent(tmp$pd.lower, .1)` and `r percent(tmp$rope.lower, .1)`, and as large as `r percent(tmp$pd.upper, .1)` and `r percent(tmp$rope.upper, .1)`, respectively.

@@

```{r}
#| label: hdist-5
#| fig.height: 2.2
#| fig.width: 4
#| fig-cap: **A**. 4,000 random draws from the posterior distribution of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). **B**. Histogram of 4,000 draws from the posterior distribution of the ratio of the valence distribution's scale over its  location $\frac{\tau_1}{\beta_1}$, and its posterior mean and 95%CI.
p1 <- post1 |> 
  ggplot(aes(b_valence1, sd_person__valence1)) +
  scale_y_continuous(
    TeX("$\\tau_1$")
  ) +
  scale_x_continuous(
    TeX("$\\beta_1$")
  ) +
  geom_point(
    shape = 1,
    col = "dodgerblue3",
    alpha = .25,
    size = .5
  )
p2 <- post1 |> 
  mutate(ratio = abs(sd_person__valence1 / b_valence1)) |> 
  ggplot(aes(ratio)) +
    scale_y_continuous(
    "Count",
    expand = expansion(0)
  ) +
  scale_x_continuous(
    TeX('$\\frac{\\tau_1}{\\beta_1}$'),
    breaks = extended_breaks()
  ) +
  stat_histinterval(
    breaks = 50,
    outline_bars = TRUE,
    slab_color = "gray10",
    slab_fill = alpha("dodgerblue1", .33),
    slab_linewidth = .2,
    justification = .01,
    point_color = "dodgerblue4", 
    interval_color = "dodgerblue4",
    .width = .90
  ) +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
  )
(p1 | p2) &
  theme(
    aspect.ratio = 1
  ) &
  plot_annotation(tag_levels = "A")
```
